CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-06-30 script/en/chapters/primes.tex
be updates

% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T W O:  P R I M E S
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section{Prime Numbers}
\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, May 1999, Updates Nov. 2000, Dec. 2001, June 2003)
\label{Label_Kapitel_2}

@@

Almost all known huge prime numbers are special candidates, called
\index{Mersenne Marin} {\em Mersenne numbers} of the form $2^p -1,$ where $p$ is
a prime. Marin Mersenne (1588-1648) was a French priest and mathematician. Not
all Mersenne numbers are prime:

$$
\begin{array}{cl}
2^2 - 1 = 3 & \Rightarrow {\rm prime} \\
2^3 - 1 = 7 & \Rightarrow {\rm prime} \\
2^5 - 1 = 31    & \Rightarrow {\rm prime} \\
2^7 - 1 = 127    & \Rightarrow {\rm prime} \\
2^{11} - 1 = 2.047 = 23 \cdot 89    & \Rightarrow  {\rm NOT~prime} !
\end{array}
$$

@@

\begin{Proof}{}
The theorem of Mersenne can be proved by contradiction%
\index{Proof by contradiction}.  We therefore assume that
there exists a composite natural number $ n $ (with real factorisation)  
$ n=n_1 \cdot n_2 $, with the property that $ 2^n -1 $ is a prime number.

@@

\hypertarget{FermatNumbers01}{}\label{FermatNumbers01}%
In the 17th century, Fermat\index{Fermat Pierre} wrote to Mersenne \index{Mersenne Marin} that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (\hyperlink{FermatNumbers02}{see below}) 
\index{Number!Fermat}\index{Fermat!number}

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 340,282,366,920,938,463,463,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  5,704,689,200,685,129,054,721
\end{eqnarray*}

@@

Comment: Numbers n that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$
but are not prime are called {\em pseudo prime numbers}
\index{Prime number!pseudo prime}\index{Number!pseudo prime}(i.e. the 
exponent is not a prime). 
The first pseudo prime number is $$ 341 = 11 \cdot 31 $$.
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Number!Carmichael} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17 $$.

A stronger test is provided by\index{Miller Gary L.}\index{Rabin Michael O.}
Miller/Rabin\footnote{
In 1976 an efficient probabilistic primality test was published by Prof. Rabin, based on a number theoretic result of Prof. Miller from the year before. \\
Prof. Miller worked at the Carnegie-Mellon University, School of Computer
Science. Prof. Rabin, born in 1931, worked at the Harvard and Hebrew University.
}%
: it is
only passed by so-called {\em strong pseudo prime numbers}. Again, there are
strong pseudo prime numbers that are not primes, but this is much less often the
case than for (simple) pseudo prime numbers. The smallest strong pseudo prime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +20 pt
\subsection{Further special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

%\begin{enumerate}

\subsubsection{Mersenne numbers $f(n) = 2^n - 1$  \quad for $ n $ prime}
%\item {\em Mersenne numbers}  $f(n) = 2^n - 1$  \quad for $ n $ prime: \\
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 2,000,000 digits are 
    known (M-38\index{Mersenne!prime number!M-38}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}

\subsubsection[Generalisation of the Mersenne numbers
   $F(k,n) = k \cdot 2^n \pm 1$]{Generalisation of the Mersenne numbers
   $F(k,n) = k \cdot 2^n \pm 1$  for $ n $ prime and $ k $ small prime}
%\item $F(k,n) = k \cdot 2^n \pm 1$  for $ n $ prime and $ k $ small prime:\\
   For this generalisation of the Mersenne 
   numbers\index{Mersenne!number!generalized}
   there are (for small $k$)
   also extremely quick prime number tests (see \cite{Knuth1981}). This can
   be performed in practice using software such as the Proths software from
   Yves Gallot\index{Gallot Yves}
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

\subsubsection{2nd generalisation of the Mersenne numbers
    $ f(n) = b^n \pm 1 $}

   This is a another possible generalisation of the Mersenne 
   numbers\index{Mersenne!number!generalized}.
   The \index{Cunningham project} \textbf{Cunningham project} determines
   the factors of all composite numbers that are formed as follows:
   $$ f(n) = b^n \pm 1  \quad {\rm for~} b = 2, 3, 5, 6, 7, 10, 11, 12 $$
   ($b$ is not equal to multiples of bases already used, such as $4, 8, 9$).

@@

%\vskip +15 pt
\hypertarget{FermatNumbers02}{}
\subsubsection[Fermat numbers]{Fermat numbers\footnotemark}
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
       \index{Number!Fermat}\index{Fermat!number}
    }
    $F(n) = 2^{2^n} + 1$

%\item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in
%       circle division.
%       As proven by Gauss\index{Gauss Carl Friedrich}, a regular $p$-edge
%       can only be constructed with the use of a pair of compasses and a 
%       ruler, when $p$ is a Fermat prime number.}
%       \index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\

       As mentioned \hyperlink{FermatNumbers01}{above} in 
       chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
       his assumption, that all numbers of this type are primes.
       Surprisingly he would have been able obtain a positive result using
       the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
F(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !
\end{array} 
$$

@@

    The Fermat number
    $$ F[2,145,351] = 2^{(2^{2,145,351})} + 1 $$ 
    is divisible by the prime
    $$ p = 3*2^{2,145,353} + 1 $$ \\
    This prime p is the largest known prime generalized Mersenne 
    number\index{Mersenne!number!generalized} and the
    5th largest known prime number today. 
    It was discovered only a few days after Michael Angel has come across the 
    largest generalized Fermat number (\hyperlink{generalizedFermatprimes}
    {see below}).  This moved GIMPS' \index{GIMPS} first prime discovery,
    M-35 = M1398269, into 7th place.

@@

\subsubsection[Verallgemeinerte Fermatzahlen]
              {Verallgemeinerte Fermatzahlen\footnotemark}
    \footnotetext{%
      The base of this power is no longer restricted to 2 . \\
      Even more generic would be:  $F(b,c,n) = b^{c^n} + 1$
    }
\hypertarget{generalizedFermatprimes}{}
\index{Fermat!number!generalized}  \\
%\vskip +15 pt
%\item {\em Generalized Fermat numbers}\footnote{%
%The base of this power is no longer restricted to 2 !}  $F(b,m) = b^{2^m} + 1$:
    Generalized Fermat numbers are more numerous than Mersenne numbers at
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

\subsubsection{Carmichael numbers\index{Numbers!Carmichael}: see above.}
%  \vskip +15 pt
%  \item Carmichael numbers: see above.

\subsubsection{Pseudo prime numbers: see above}
%  \item Pseudo prime numbers: see above.

\subsubsection{Strong pseudo prime numbers: see above}
%  \item Strong pseudo prime numbers: see above.

\subsubsection[Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$]
{Idea based on Euclid's proof (infinite many prime numbers) $p_1 \cdot p_2 \cdots p_n +1$}
%  \item Idea based on \hyperlink{euclid}{Euclid's proof} (infinite many prime
% numbers) $p_1 \cdot p_2 \cdots p_n +1$:

$$
\begin{array}{lll}
2{\cdot}3 +1 &      = 7 &          \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5 +1 &      = 31    &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5{\cdot}7 +1 &      = 211   &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdots}11 +1 &        = 2,311  &      \mapsto {\rm ~prime} \\
2\cdot3 \cdots 13 +1 &  = 59 \cdot 509 &    \mapsto {\rm ~NOT~prime} ! \\
2\cdot3 \cdots 17 +1 &  = 19 \cdot 97 \cdot 277 &   \mapsto {\rm ~NOT~prime} !
\\
\end{array}
$$

\subsubsection{As above but except $+1$: $p_1 \cdot p_2 \cdots p_n -1$}
%  \item As above but except $+1$: $p_1 \cdot p_2 \cdots p_n -1$

$$
\begin{array}{lll}
2\cdot 3 -1     &   = 5 &   \mapsto {\rm ~prime} \\
2\cdot 3 \cdot  5  -1   &   = 29 &  \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 7  -1   &   = 11 \cdot 19 & \mapsto {\rm ~NOT~prime} ! \\
2\cdot 3 \cdots 11 -1   &   = 2,309 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 13 -1   &   = 30,029 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 17 -1    &  = 61 \cdot 8,369 &   \mapsto {\rm ~NOT~prime!}
\end{array} 
$$

\subsubsection[Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$]
    {Euclidean numbers $e_n = e_0 
    \cdot e_1 \cdots e_{n-1} + 1$  with $n$ greater than or equal to 
    $1$ and $e_0 := 1$}\index{Euclidean number}
%  \item \index{Euclidean number} {\em Euclidean numbers} $e_n = e_0 
%    \cdot e_1 \cdots e_{n-1} + 1$  with $n$ greater than or equal to 
%    $1$ and $e_0 := 1$. \\
    $e_{n-1}$ is not the $(n-1)$th prime number, but the number 
    previously found here.
    Unfortunately this formula is not open but recursive.
    The sequence starts with

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore\index{Prime number!relative prime}\index{Relatively prime} {\em relatively prime}.

\subsubsection{$f(n) = n^2 + n + 41$}
%  \item $f(n) = n^2 + n + 41$:\\
  This sequence starts off very {\em promisingly},  
  but is far from being a proof.

@@

    This means there is no hope in looking for further formulae 
    similar to that of type 11. or 12.

\subsubsection[Catalan's conjecture]{Catalan's conjecture\footnotemark}
    \footnotetext{%
Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
After him the so-called {\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
$=  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.
    }
Catalan \index{Catalan Eugene}
%\item \index{Catalan Eugene} \index{Catalan number} Eugene Cahrles 
%  Catalan\footnote{%
%Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
%After him the so-called {\em Catalan numbers}
%$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
%$=  1, 2, 5, 14, 42, 132, 429, 1430, 
%4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
%are named.
%}
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

\vskip +3 pt
\begin{theorem}\label{thm-pz-density}
For large n: The value of the $n$-th prime $P(n)$ is asymptotic to $n \cdot
ln(n)$, i.e. the limit of the relation  $P(n)/(n\cdot \ln n)$ is equal to $1$ if
$n$ tends to infinity.
\end{theorem}

For large $n$, $P(n)$  lies between  $2n$  and  $n^2$. This means that there are
fewer prime numbers than even natural numbers but more prime numbers than
square numbers.
This already is true from the 5th prime, as you can see in 
\hyperlink{ntePrimzahl}{the table \ref{s:ntePrimzahl}}: $P(5) = 11$.

The definition is similar for the number of prime numbers $PI(x)$
that do not exceed the maximum value $x$:

@@

For a binary number (number in the binary system) x of the length $250$ bits
($2^{250}$ is approximately = $1.809 251 * 10^{75}$): 
$$ PI(x) = 2^{250} / (250 \cdot \ln 2) \; {\rm is~approximately} \; =  2^{250} /173.28677 = 1.045 810 \cdot 10^{73}. $$
We can therefore expect that the set of numbers with a bit length of less than
250 contains approximately $10^{73}$ primes (a reassuring result?!).

>>>>

% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T W O :  P R I M E S
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\hypertarget{Kapitel_2}{}

\section{Prime Numbers}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999, Updates Nov. 2000, Dec. 2001, June 2003)

@@

Almost all known huge prime numbers are special candidates, called
\index{Mersenne Marin} {\em Mersenne numbers}\footnote{%
Marin Mersenne, French priest and mathematician, Sep 08, 1588 -- Sep 01, 1648.
\index{Mersenne Marin}
}
of the form $2^p -1,$ where $p$ is
a prime. Not all Mersenne numbers are prime:

$$
\begin{array}{cl}
2^2 - 1 = 3 & \Rightarrow {\rm prime} \\
2^3 - 1 = 7 & \Rightarrow {\rm prime} \\
2^5 - 1 = 31    & \Rightarrow {\rm prime} \\
2^7 - 1 = 127    & \Rightarrow {\rm prime} \\
2^{11} - 1 = 2,047 = 23 \cdot 89    & \Rightarrow  {\rm NOT~prime} !
\end{array}
$$

@@

\begin{Proof}{}
The theorem of Mersenne can be proved by contradiction%
\index{Proof by contradiction}.  We therefore assume that
there exists a composite natural number $ n $ (with real factorisation)  
$ n=n_1 \cdot n_2 $
, with the property that $ 2^n -1 $ is a prime number.

@@

\hypertarget{FermatNumbers01}{}\label{FermatNumbers01}%
In the 17th century, Fermat\footnote{%
Pierre de Fermat, French mathematician, Aug 17, 1601 -- Jan 12, 1665.
\index{Fermat Pierre}
}
\index{Fermat Pierre} wrote to Mersenne \index{Mersenne Marin} that he 
presumed that all numbers of the form 
$$ f(n) = 2^{2^n} + 1 $$ 
are prime for all whole numbers $ n \geq 0$ 
(\hyperlink{FermatNumbers02}{see below}, chapter~\ref{L-FermatNumbers02}).
\index{Number!Fermat}\index{Fermat!number}

As early as in the 19th century, it was discovered that the $29$-digit number $$
f(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
f(7) & = & 340,282,366,920,938,463,463,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  5,704,689,200,685,129,054,721
\end{eqnarray*}\label{f7}

@@

\vskip +15 pt
\paragraph{Pseudo prime numbers}%
\index{Prime number!pseudo prime} \index{Number!pseudo prime}%
\hypertarget{HT-Pseudoprimenumber01}{}\label{L-Pseudoprimenumber01}%
\mbox{}
\vskip +10 pt
Numbers n that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$
but are not prime are called {\em pseudo prime numbers}
(i.e. the exponent is not a prime). 
The first pseudo prime number is $$ 341 = 11 \cdot 31 .$$

\vskip +15 pt
\paragraph{Carmichael numbers}%
\index{Number!Carmichael}%
\hypertarget{HT-Carmichael-number01}{}\label{L-Carmichael-number01}%
\mbox{}
\vskip +10 pt
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called {\em Carmichael numbers}. The first of these is
$$ 561 = 3 \cdot 11 \cdot 17 .$$

\vskip +15 pt
\paragraph{Strong pseudo prime numbers}%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
\hypertarget{HT-Strongpseudoprimenumber01}{}\label{L-Strongpseudoprimenumber01}%
\mbox{}
\vskip +10 pt
A stronger test is provided by\index{Miller Gary L.}\index{Rabin Michael O.}
Miller/Rabin\footnote{
In 1976 an efficient probabilistic primality test was published by Prof. Rabin, based on a number theoretic result of Prof. Miller from the year before. \\
Prof. Miller worked at the Carnegie-Mellon University, School of Computer
Science. Prof. Rabin, born in 1931, worked at the Harvard and Hebrew University.
}%
: it is only passed by so-called {\em strong pseudo prime numbers}. 
Again, there are strong pseudo prime numbers that are not primes, but this is
much less often the case than for (simple) pseudo prime numbers or for 
Carmichael numbers. The smallest strong pseudo prime number base $2$ is 
$$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +20 pt
\subsection{Overview special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

% --------------------------------------------------------------------------
\subsubsection{Mersenne numbers $f(n) = 2^n - 1$  \quad for $ n $ prime}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 2,000,000 digits are 
    known (M-38\index{Mersenne!prime number!M-38}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}

% --------------------------------------------------------------------------
\subsubsection
   [Generalized Mersenne numbers $f(k,n) = k \cdot 2^n \pm 1$]
   {Generalized Mersenne numbers
   $f(k,n) = k \cdot 2^n \pm 1 $  for $ n $ prime and $ k $ small prime}
   For this first generalisation of the Mersenne 
   numbers\index{Mersenne!number!generalized}
   there are (for small $k$)
   also extremely quick prime number tests (see \cite{Knuth1981}). This can
   be performed in practice using software such as the Proths software from
   Yves Gallot\index{Gallot Yves}
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

% --------------------------------------------------------------------------
\subsubsection
   [Generalized Mersenne numbers $ f(b,n) = b^n \pm 1 $ / Cunningham project]
   {Generalized Mersenne numbers 
   $ f(b,n) = b^n \pm 1 $ / The Cunningham project}

   This is a another possible generalisation of the Mersenne 
   numbers\index{Mersenne!number!generalized}.
   The \index{Cunningham project} \textbf{Cunningham project} determines
   the factors of all composite numbers that are formed as follows:
   $$ f(b,n) = b^n \pm 1  \quad {\rm for~} b = 2, 3, 5, 6, 7, 10, 11, 12 $$
   ($b$ is not equal to multiples of bases already used, such as $4, 8, 9$).

@@

% --------------------------------------------------------------------------
\hypertarget{FermatNumbers02}{}
\subsubsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Number!Fermat}\index{Fermat!number}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    Surprisingly he would have been able obtain a positive result using
    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{f7})}  
																 & \mapsto {\rm ~NOT~prime} !

\end{array} 
$$

@@

    The Fermat number
    $$ f(2,145,351) = 2^{(2^{2,145,351})} + 1 $$ 
    is divisible by the prime
    $$ p = 3*2^{2,145,353} + 1 $$ \\
    This prime p is the largest known prime generalized Mersenne 
    number\index{Mersenne!number!generalized} and the
    5th largest known prime number today. 
    It was discovered only a few days after Michael Angel has come across the 
    largest generalized Fermat number (\hyperlink{generalizedFermatprimes}
    {see below}).  This moved GIMPS' \index{GIMPS} first prime discovery,
    M-35 = M1398269, into 7th place.

@@

% --------------------------------------------------------------------------
\subsubsection[Generalized Fermat numbers $f(b,n) = b^{2^n} + 1$]
              {Generalized Fermat numbers\footnotemark~$f(b,n) = b^{2^n} + 1$}
    \footnotetext{%
      The base of this power is no longer restricted to 2 . \\
      Even more generic would be:  $f(b,c,n) = b^{c^n} \pm 1$
    }
\hypertarget{generalizedFermatprimes}{}
\index{Fermat!number!generalized} 
    Generalized Fermat numbers are more numerous than Mersenne numbers at
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

% --------------------------------------------------------------------------
\subsubsection{Carmichael numbers\index{Number!Carmichael}}
As mentioned \hyperlink{HT-Carmichael-number01}{above} in
chapter~\ref{L-Carmichael-number01} not all Carmichael numbers are prime.

% --------------------------------------------------------------------------
\subsubsection{Pseudo prime numbers%
\index{Prime number!pseudo prime}\index{Number!pseudo prime}%
}
See \hyperlink{HT-Pseudoprimenumber01}{above} in
chapter~\ref{L-Pseudoprimenumber01}.

% --------------------------------------------------------------------------
\subsubsection{Strong pseudo prime numbers%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
}
See \hyperlink{HT-Strongpseudoprimenumber01}{above} in
chapter~\ref{L-Strongpseudoprimenumber01}.

% --------------------------------------------------------------------------
\subsubsection
    [Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$]
    {Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$}
This idea is based on \hyperlink{euclid}{Euclid's proof} that there are
infinite many prime numbers.
$$
\begin{array}{lll}
2{\cdot}3 +1 &      = 7 &          \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5 +1 &      = 31    &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5{\cdot}7 +1 &      = 211   &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdots}11 +1 &        = 2,311  &      \mapsto {\rm ~prime} \\
2\cdot3 \cdots 13 +1 &  = 59 \cdot 509 &    \mapsto {\rm ~NOT~prime} ! \\
2\cdot3 \cdots 17 +1 &  = 19 \cdot 97 \cdot 277 &   \mapsto {\rm ~NOT~prime} !
\\
\end{array}
$$

% --------------------------------------------------------------------------
\subsubsection{As above but $-1$ except $+1$: $p_1 \cdot p_2 \cdots p_n -1$}
$$
\begin{array}{lll}
2\cdot 3 -1     &   = 5 &   \mapsto {\rm ~prime} \\
2\cdot 3 \cdot  5  -1   &   = 29 &  \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 7  -1   &   = 11 \cdot 19 & \mapsto {\rm ~NOT~prime} ! \\
2\cdot 3 \cdots 11 -1   &   = 2,309 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 13 -1   &   = 30,029 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 17 -1    &  = 61 \cdot 8,369 &   \mapsto {\rm ~NOT~prime!}
\end{array} 
$$

% --------------------------------------------------------------------------
\subsubsection[Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$]
              {Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$
	       with $ n \geq 1 $ and $ e_0 := 1 $}
%	       with $n$ greater than or equal to $1$ and $e_0 := 1$}
    \index{Euclidean number}
    $e_{n-1}$ is not the $(n-1)$th prime number, but the number 
    previously found here.
    Unfortunately this formula is not open but recursive.
    The sequence starts with

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful.

Remark: However, what is special about these numbers is that any pair of 
them does not have a common factor other than $1$. They are therefore%
\index{Prime number!relative prime}\index{Relatively prime}
{\em relatively prime}.

\pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% --------------------------------------------------------------------------
\subsubsection{$f(n) = n^2 + n + 41$}
\label{L-Polynomfunktion01-41}
   This sequence starts off very {\em promisingly},  
   but is far from being a proof.

@@

    This means there is no hope in looking for further formulae (functions) 
    similar to that in chap.~\ref{L-Polynomfunktion01-41} or 
    chap.~\ref{L-Polynomfunktion02-1601}.

% --------------------------------------------------------------------------
\subsubsection[Catalan's conjecture]{Catalan's conjecture\footnotemark}
    \footnotetext{%
Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
After him the so-called {\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
$=  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.
    }
Catalan \index{Catalan Eugene}  \index{Number!Catalan}
conjectured that $ C_4 $ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

\vskip +15 pt
\paragraph{The value of the $n$-th prime $P(n)$}%
\index{P(n)}%
\mbox{}
\vskip +10 pt

\begin{theorem}\label{thm-pz-density}
For large n: The value of the $n$-th prime $P(n)$ is asymptotic to $n \cdot
ln(n)$, i.e. the limit of the relation  $P(n)/(n\cdot \ln n)$ is equal to $1$ if
$n$ tends to infinity.
\end{theorem}

For large $n$, $P(n)$  lies between  $2n$  and  $n^2$. This means that there are
fewer prime numbers than even natural numbers but more prime numbers than
square numbers.

This already is true from the 5th prime, as you can see in 
\hyperlink{ntePrimzahl}{the table \ref{s:ntePrimzahl}}: $P(5) = 11$.

\vskip +15 pt
\paragraph{The number of prime numbers $PI(x)$}%
\index{PI(x)}%
\mbox{}
\vskip +10 pt
The definition is similar for the number of prime numbers $PI(x)$
that do not exceed the maximum value $x$:

@@

For a binary number\footnote{%
Number written in the binary system consists only of the digits 0 and 1.
}  
$ x $ of the length of $250$ bits
($2^{250}$ is approximately = $1.809 251 * 10^{75}$)
it is: 
$$ PI(x) = 2^{250} / (250 \cdot \ln 2)  \;\;  {\rm is~approximately}
\; = 2^{250} / 173.28677 = 1.045 810 \cdot 10^{73}. $$
We can therefore expect that the set of numbers with a bit length of less than
250 contains approximately $10^{73}$ primes (a reassuring result?!).
